---
title: "Document S2"
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    number-sections: false
    toc: false
    fig-pos: H
    tbl-cap-location: top
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{float}
        \usepackage{pdflscape}
        \usepackage[margin=2.2cm]{geometry}
        \setlength{\LTpre}{0pt}
        \setlength{\LTpost}{6pt}
  docx:
    toc: false
execute:
  echo: false
  warning: false
  message: false
  freeze: false
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
library(readxl)

options(knitr.kable.NA = "--")

trait_labels <- c(fapar = "FAPAR", fcover = "FCOVER", laie = "LAIe")

result_files <- Sys.glob("results/evaluation/*/summary_metrics.csv")
stopifnot(length(result_files) == 3)

results <- bind_rows(lapply(result_files, function(path) {
  trait <- basename(dirname(path))
  read.csv(path, check.names = FALSE) |>
    mutate(trait = unname(trait_labels[trait]))
})) |>
  mutate(
    improved = delta_rmse < 0,
    validation_group = sub(
      "Temperate deciduous & mixed forests",
      "Temperate broadleaf & mixed forests",
      validation_group,
      fixed = TRUE
    )
  )

stopifnot(
  max(abs(results$delta_rmse - results$delta_rmse_recomputed)) < 1e-10,
  max(abs(results$delta_rmse - results$delta_rmse_measurement_recomputed)) < 1e-10
)

recoding_raw <- read_excel(
  "temp_crosstab_recoded_with_counts.xlsx",
  sheet = "manual_recoding_logic",
  col_names = FALSE
)

# Rows 22--29 contain the final group names. Rows 11--17 contain the
# biome-by-NLCD recoding matrix. Retain one row for every included combination.
group_names <- recoding_raw[22:29, 2:3] |>
  setNames(c("Validation group", "Group")) |>
  mutate(Group = as.integer(Group))

biome_names <- as.character(unlist(recoding_raw[11, 2:10]))
nlcd_names <- as.character(unlist(recoding_raw[12:17, 1]))
recoding_matrix <- as.matrix(recoding_raw[12:17, 2:10])

recoding_entries <- bind_rows(lapply(seq_along(nlcd_names), function(i) {
  bind_rows(lapply(seq_along(biome_names), function(j) {
    data.frame(
      Group = suppressWarnings(as.integer(recoding_matrix[i, j])),
      Biomes = biome_names[j],
      `NLCD land covers` = nlcd_names[i],
      check.names = FALSE
    )
  }))
})) |>
  filter(Group %in% 1:8) |>
  left_join(group_names, by = "Group") |>
  arrange(Group, Biomes, `NLCD land covers`)

stopifnot(nrow(recoding_entries) == 26L)

# Within each validation group, combine biomes that share the same set of
# included land-cover classes. This preserves the exact mapping in one row.
recoding <- recoding_entries |>
  group_by(Group, `Validation group`, Biomes) |>
  summarise(
    land_covers = paste(unique(`NLCD land covers`), collapse = ", "),
    .groups = "drop"
  ) |>
  group_by(Group, `Validation group`, land_covers) |>
  summarise(
    biomes = paste(Biomes, collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(Group, biomes) |>
  mutate(mapping = paste0(biomes, ": *", land_covers, "*")) |>
  group_by(Group, `Validation group`) |>
  summarise(
    `Included biome--land-cover combinations` = paste(mapping, collapse = "; "),
    .groups = "drop"
  ) |>
  arrange(Group) |>
  select(-Group)

stopifnot(nrow(recoding) == 8L)
```

\setcounter{figure}{9}
\setcounter{table}{9}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

# Note S1. Biome- and vegetation-type-specific model evaluation

**Methods.** To assess whether vegetation-specific calibration could improve retrieval performance relative to a single global model (henceforth _generic_ model), 
we repeated model calibration for groups defined by combinations of biome and vegetation type. 
Rather than defining groups by biome or land cover alone, we grouped biome × land-cover combinations representing similar vegetation structure and environmental conditions, while ensuring sufficient data for model calibration and evaluation.
Biome assignments followed the RESOLVE ecoregion biome labels. Land-cover classes were derived from the NLCD labels associated with each GROUNDED-EO reference measurement (RM). We simplified the original NLCD classification into six vegetation classes: cropland, deciduous/mixed forest, evergreen forest, grassland, shrubland, and wetland.
We then cross-tabulated these vegetation classes with biome assignments and combined reference measurements into eight groups (@tbl-validation-groups). 
Combinations that could not be assigned to a sufficiently represented group were excluded, removing 342 RMs. The resulting grouped dataset contained 10,696 RMs across 44 ecoregions.

We repeated the model calibration described in the main Methods independently for each biome–vegetation group and for a generic model using all RMs assigned to the retained groups. 
The same procedure and parameter search space were used for all models and traits. 
All models used the S2BIOPHYS hybrid PROSAIL inversion with a multilayer perceptron. Predictors comprised ten Sentinel-2 reflectance bands and three observation-geometry angles.
To reduce computational costs, we used a reduced optimization budget of 1,500 trials per model, comprising 600 random startup trials followed by 900 trials using the tree-structured Parzen estimator (TPE).
The generic model was calibrated separately for this experiment using the same reduced optimization budget.

Instead of the five-fold ecoregion-blocked cross-validation used for the main S2BIOPHYS model, we used three folds to accommodate the smaller number of ecoregions within individual biome–vegetation groups. All RMs from the same ecoregion were retained within the same fold.

For the primary comparison, we used stacked out-of-sample (OOS) predictions, with each RM predicted only when its ecoregion was held out. Vegetation-specific and generic models were evaluated on identical RMs within each group using RMSE and $R^2$. The group-specific $R^2$ used the mean of each validation group. The global-reference $R^2_g$ used the mean across all groups. We defined $\Delta\mathrm{RMSE}=\mathrm{RMSE}_{\mathrm{specific}}-\mathrm{RMSE}_{\mathrm{generic}}$, with negative values indicating better vegetation-specific performance.

We quantified uncertainty in $\Delta$RMSE by paired cluster bootstrap resampling of the existing OOS predictions. We sampled ecoregions with replacement and retained all RMs from each sampled ecoregion. The same resampled clusters were used for both models. We used 5,000 replicates and percentile-based 95% confidence intervals. Intervals were reported only for validation groups with at least 10 ecoregions. Groups with fewer ecoregions retained point estimates because their spatial replication was insufficient for reliable bootstrap intervals.

As a sensitivity analysis, we also used a paired bootstrap of individual RMs. We sampled RMs with replacement within each validation group and across the pooled dataset. The same sampled RMs were used for both models. We used 5,000 replicates and percentile-based 95% confidence intervals. These intervals treat RMs as independent and do not account for spatial dependence within ecoregions.

**Results.** Overall, the generic model performed better when pooling RMs across all validation groups, 
with lower pooled RMSE for all three traits (@fig-rmse-change; @tbl-performance). 
Vegetation-specific calibration provided improvements primarily in well-represented groups. 
For LAIe, specific models reduced RMSE in the three largest groups, 
whereas performance was generally less consistent in groups with fewer RMs. 
This indicates that vegetation-specific calibration can improve retrievals when 
sufficient reference data are available, but that pooling data across vegetation types provides more robust performance overall. 
Bootstrap intervals were therefore estimated for the three temperate groups and for the pooled comparison. Groups with four or five ecoregions are shown without intervals because spatial replication was insufficient.
All reported 95% confidence intervals included zero.
The individual-RM bootstrap produced intervals for every group (@fig-rmse-change-measurements). These intervals are descriptive because they do not preserve the ecoregion-level dependence structure.

```{r}
#| label: fig-rmse-change
#| fig-cap: "Change in out-of-sample RMSE for biome--vegetation-specific models relative to the generic model. Horizontal lines are paired ecoregion-bootstrap 95% confidence intervals and are shown only for groups with at least 10 ecoregions. Negative values indicate lower error for the specific model. Point size represents the number of reference measurements."
#| fig-width: 7.4
#| fig-height: 3.5
plot_data <- results |>
  mutate(
    validation_group = factor(
      validation_group,
      levels = c(
        "Pooled across validation groups",
        rev(unique(results$validation_group[
          results$trait == "FAPAR" &
            results$validation_group != "Pooled across validation groups"
        ]))
      )
    )
  )

ggplot(plot_data, aes(delta_rmse, validation_group, size = N, colour = improved)) +
  geom_vline(xintercept = 0, linewidth = 0.35, colour = "grey45") +
  geom_point(alpha = 0.9) +
  facet_wrap(~trait, scales = "free_x", nrow = 1) +
  scale_colour_manual(
    values = c(`TRUE` = "#0072B2", `FALSE` = "#D55E00"),
    labels = c(`TRUE` = "Specific better", `FALSE` = "Generic better")
  ) +
  scale_size_area(max_size = 5, breaks = c(500, 1000, 2000, 3000)) +
  labs(x = expression(Delta*"RMSE (specific - generic)"), y = NULL, colour = NULL, size = "Reference\nmeasurements") +
  theme_bw(base_size = 9) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey94"),
    strip.text = element_text(face = "bold")
  )
```

```{r}
#| label: fig-rmse-change-measurements
#| fig-cap: "Change in out-of-sample RMSE for biome--vegetation-specific models relative to the generic model. Horizontal lines are paired individual-reference-measurement bootstrap 95% confidence intervals. Negative values indicate lower error for the specific model. Point size represents the number of reference measurements. These intervals do not account for dependence among measurements from the same ecoregion."
#| fig-width: 7.4
#| fig-height: 3.5
ggplot(plot_data, aes(delta_rmse, validation_group, size = N, colour = improved)) +
  geom_vline(xintercept = 0, linewidth = 0.35, colour = "grey45") +
  geom_point(alpha = 0.9) +
  facet_wrap(~trait, scales = "free_x", nrow = 1) +
  scale_colour_manual(
    values = c(`TRUE` = "#0072B2", `FALSE` = "#D55E00"),
    labels = c(`TRUE` = "Specific better", `FALSE` = "Generic better")
  ) +
  scale_size_area(max_size = 5, breaks = c(500, 1000, 2000, 3000)) +
  labs(x = expression(Delta*"RMSE (specific - generic)"), y = NULL, colour = NULL, size = "Reference\nmeasurements") +
  theme_bw(base_size = 9) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey94"),
    strip.text = element_text(face = "bold")
  )
```

```{r}
#| label: word-break-before-table-s10
#| results: asis
#| echo: false
if (knitr::pandoc_to() == "docx") {
  cat('```{=openxml}\n<w:p><w:r><w:br w:type="page"/></w:r></w:p>\n```\n')
}
```

```{r}
#| label: tbl-validation-groups
#| tbl-cap: "Biome and NLCD land-cover combinations assigned to the eight validation groups. Biome names are followed by the included NLCD land-cover classes in italics; semicolons separate distinct mappings."
#| results: asis
if (knitr::is_latex_output()) {
  escape_ampersands <- function(x) {
    gsub("&", paste0(intToUtf8(92), "&"), x, fixed = TRUE)
  }
  recoding_latex <- recoding
  recoding_latex[[1]] <- escape_ampersands(recoding_latex[[1]])
  recoding_latex[[2]] <- vapply(
    strsplit(recoding[[2]], "; ", fixed = TRUE),
    function(parts) {
      formatted <- vapply(parts, function(part) {
        pieces <- strsplit(part, ": ", fixed = TRUE)[[1]]
        paste0(
          escape_ampersands(pieces[1]),
          ": \\textit{",
          escape_ampersands(gsub("\\*", "", pieces[2])),
          "}"
        )
      }, character(1))
      paste(formatted, collapse = "; ")
    },
    character(1)
  )
  cat("\\begingroup\\footnotesize\n")
  print(kable(
    recoding_latex,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    escape = FALSE,
    align = c("p{3.2cm}", "p{8.2cm}")
  ))
  cat("\n\\endgroup\n")
} else {
  print(kable(recoding, format = "pipe", align = c("l", "l")))
}
```

```{r}
#| label: word-break-before-table-s11
#| results: asis
#| echo: false
if (knitr::pandoc_to() == "docx") {
  cat('```{=openxml}\n<w:p><w:r><w:br w:type="page"/></w:r></w:p>\n```\n')
}
```

```{r}
#| label: tbl-performance
#| tbl-cap: "Ecoregion-blocked out-of-sample performance of biome--vegetation-specific and generic models. $R^2_g$ uses the mean across all groups. $\\Delta$RMSE is specific minus generic; negative values indicate lower error for the specific model. Paired ecoregion-bootstrap 95% confidence intervals are reported for groups with at least 10 ecoregions; dashes indicate insufficient spatial replication."
#| results: asis
performance_table <- results |>
  transmute(
    Trait = trait,
    `Validation group` = gsub("&", "\\\\&", validation_group, fixed = TRUE),
    N = N,
    `RMSE specific` = sprintf("%.3f", rmse_specific),
    `RMSE generic` = sprintf("%.3f", rmse_generic),
    `$\\Delta$RMSE` = sprintf("%+.3f", delta_rmse),
    `$R^2$ specific` = sprintf("%.3f", r2_specific),
    `$R^2$ generic` = sprintf("%.3f", r2_generic),
    `$R^2_g$ specific` = ifelse(is.na(r2_global_specific), "--", sprintf("%.3f", r2_global_specific)),
    `$R^2_g$ generic` = ifelse(is.na(r2_global_generic), "--", sprintf("%.3f", r2_global_generic))
  )

trait_breaks <- rep("", nrow(performance_table))
trait_breaks[c(9, 18)] <- "\\addlinespace[2pt]\\midrule\\addlinespace[2pt]"

if (knitr::is_latex_output()) {
  cat("\\begin{landscape}\n\\begingroup\\scriptsize\n")
  print(kable(
    performance_table,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    escape = FALSE,
    linesep = trait_breaks,
    align = c("l", "p{3.3cm}", rep("r", 9))
  ))
  cat("\n\\endgroup\n\\end{landscape}\n")
} else {
  performance_word <- results |>
    transmute(
      Trait = trait,
      `Validation group (N)` = paste0(validation_group, " (", N, ")"),
      `Specific (RMSE / R2 / R2g)` = sprintf(
        "%.3f / %.3f / %s",
        rmse_specific,
        r2_specific,
        ifelse(is.na(r2_global_specific), "--", sprintf("%.3f", r2_global_specific))
      ),
      `Generic (RMSE / R2 / R2g)` = sprintf(
        "%.3f / %.3f / %s",
        rmse_generic,
        r2_generic,
        ifelse(is.na(r2_global_generic), "--", sprintf("%.3f", r2_global_generic))
      ),
    )

  cat("\n**FAPAR**\n\n")
  print(kable(
    filter(performance_word, Trait == "FAPAR") |> select(-Trait),
    format = "pipe",
    align = c("l", "c", "c", "r")
  ))
}
```

```{r}
#| label: word-performance-fcover
#| results: asis
if (knitr::pandoc_to() == "docx") {
  cat('```{=openxml}\n<w:p><w:r><w:br w:type="page"/></w:r></w:p>\n```\n')
  cat("\n**FCOVER**\n\n")
  print(kable(
    filter(performance_word, Trait == "FCOVER") |> select(-Trait),
    format = "pipe",
    align = c("l", "c", "c", "r")
  ))
}
```

```{r}
#| label: word-performance-laie
#| results: asis
if (knitr::pandoc_to() == "docx") {
  cat('```{=openxml}\n<w:p><w:r><w:br w:type="page"/></w:r></w:p>\n```\n')
  cat("\n**LAIe**\n\n")
  print(kable(
    filter(performance_word, Trait == "LAIe") |> select(-Trait),
    format = "pipe",
    align = c("l", "c", "c", "r")
  ))
}
```
